Ricci-Ollivier curvature of two random walks on rooted phylogenetic subtree-prune-regraft graph

### 2015-03-06 curvature.tex
started drafting computational parts of section 3

>>>>

\subsection{Computing curvature values}
We computed all of the curvatures for pairs of trees with up to seven leaves to gain an intuitive understanding of the distribution of curvatures.
We first used \dots to compute the rSPR treespaces for four to seven leaves.
This required \dots time.
We then used \dots \todo{ILP, citation, github reference} to compute curvatures for given pairs of trees.
This would have required an enormous amount of computation to directly compute curvatures for the $((2n-3)!!)^2$ pairs of trees with $n$ leaves.
We instead exploited the fact that pairs of trees which are equivalent modulo label renumbering are symmetric in the rSPR graph and therefore guaranteed to have the same curvature.
We thus directly computed curvature values for one representative pair from each such equivalance class, or tanglegram. \todo{cite} \todo{indicate that a paper on tanglegrams is in preparation}

\subsection{Simulating random walks on the rSPR graph}
We next explored the correspondence between curvature and graph effects.
To do so, we first simulated Metropolis-Hastings random walks on the rSPR graphs with up to seven leaves.
The uniform random walk moves to a neighboring vertex in the graph uniformly at random.
However, this makes the walk more likely to sample higher degree vertices.
In contrast, the Metropolis Hastings (MH) random walk proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, $\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
The MH random walk is guaranteed to sample each tree uniformly at random and is therefore representative of a phylogenetic MCMC program sampling trees under a uniform prior.

To efficiently simulate the MH random walk, we developed a linear-time algorithm for proposing SPR moves that does not require the rSPR graph to be precomputed.
Naively, this would require $\OhOf{n^3}$ time---$\OhOf{n}$ time to generate each of the $\OhOf{n^2}$ neighbors of a given tree and then pick one uniformly at random.
To eliminate an $\OhOf{n^2}$ factor, we developed a deterministic ordering of SPR moves such that an arbitrary SPR neighbor can be selected by its index in $\OhOf{n}$ time.
Song \todo{cite} developed a recursive formula for the degree of a tree and an explicit equation that works in terms of the distribution of heights of each node.
This formula does not group SPR moves that move a particularly subtree, and thus would still require $\OhOf{n^2}$ time to select a specific SPR neighbor by index.

Instead, we consider the distribution of SPR moves in terms of the number of nodes contained within a subtree.
Recall that a tree with $n$ leaves has $2n-1$ nodes.
Given a subtree $R$ with $x$ nodes, observe that there are $2n-1 - x$ possible locations to regraft $R$.
However, some of these moves will result in the same neighboring tree as other SPR moves.
In particular:
\begin{enumerate}
\item Moving $R$ to its sibling results in the same tree, not a neighboring tree,
\item Moving $R$ to its parent results in the same tree,
\item Moving $R$ to its grandparent is the same as moving its aunt to its sibling
\item Moving $R$ to its aunt is the same as moving its aunt to R.
\end{enumerate}
We can thus assign $2n-1-x-4$ moves to any node at a depth greater than 1 and $2n-1-x-2$ moves to depth 1 nodes (lacking both an aunt and grandparent).
In this manner, we achieve an alternative solution for computing the degree of a tree:
\todo{equation}
In particular, this formulation groups moves that move the same subtree.
We can thus apply the following algorithm to select a neighbor uniformly at random:

\begin{enumerate}
\item	compute the degree
\item pick a random number $r$ from $1$ to the degree
\item uniquely order the tree by the smallest alphanumeric label of descendants
\item preorder the nodes of $R$
\item for each tree node $R$ in order and while $r > 0$
	\begin{enumerate}
		\item decrease $r$ by $2n-1-x-4$ or $2n-1-x-2$
		\item if $r < 0$, pick the $\abs{r}$ valid move of $R$ in order
	\end{enumerate}
\item apply the corresponding move and return the corresponding tree
\end{enumerate}

This requires linear time and space.

\subsection{Mean access time computation}

We then computed mean access times (MAT) between pairs of trees in the MH random walks---the mean number of iterations required to move from one tree to the other. \todo{cite}
MAT values have been productively applied to demonstrate graph effects in uSPR subgraphs of real MCMC samples using MrBayes . \todo{cite sprmix}
We used the same sprspace software to compute MAT values for our simulated random walks.

\subsection{Results}
